---
title: "Bacterial cell counts"
author: "Fabian Roger"
date: "21 Jun 2016"
output:
  html_document:
    fig_caption: yes
    toc: yes
  pdf_document:
    latex_engine: lualatex
    toc: yes
---

In this script I import and clean the data from the 
Carbon Source Utilization Profiling using the BIOLOG™ Ecoplates



__this script imports:__ 

+ Ecolog/BacCounts.txt      # raw Flow Cyto data
+ BOT_ID.txt                # sample metadata
   
__this script does:__

  + quality control 
  + bias correction
  
__this script exports:__
  
  data frames:
  
  
```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message=FALSE)
```

```{r load packages, echo = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)

```


```{r, load data}
Bac <- read.table("BacCounts.txt" , header = T , sep = "\t")
ID <- read.table("ID.txt")

ID <- mutate(ID, bot_label = paste(tolower(Sampling), formatC(BOT, width = 2, flag = 0), sep = "-"))

```


! for Tox 0 the Botlle number doesn't seem to correspond to the right cell numbers. Check! 

(see e.g. ` r Bac %>% filter(sampling == ymd("2013-10-20"), DIV == 1, Tox == 0) %>% arrange(sampling, Cells, BOT))` to see what I mean)

```{r}
# transform $sampling into POSIXct date
Bac$sampling <- dmy(Bac$sampling)

# add with / without tox level
Bac <- mutate(Bac, ToxYN = ifelse(sampling < ymd(20131010), "NO", "YES"))


# melt to long format

Bac <- gather(Bac, Cells, Counts, CellsL, CellsD, CellsT)

Bac$Cells <- factor(Bac$Cells, levels=c("CellsT","CellsL","CellsD"))

# plot Cell counts by diversity and Tox

Bac %>% filter(DIV != "S") %>% 
  ggplot(., aes(x = sampling, y=Counts, colour=Cells))+
  facet_wrap(~DIV * ToxC,ncol=7, labeller = label_wrap_gen(multi_line = F))+
  geom_point(position = position_jitter(width = 0.2))+
  geom_vline(xintercept = as.numeric(ymd("2013-10-10")), linetype = "dashed")+
  stat_summary(geom = "line", fun.y = "mean")+
  theme_bw()+
  scale_colour_manual(values=c("darkgreen","darkred","darkblue"))+
  theme(axis.text.x=element_text(angle=45,hjust = 1, vjust = 1 ))+
  scale_x_date(breaks = unique(Bac$sampling)[seq(1 , length(unique(Bac$sampling)), 2)])

```

plot cell counts relative to 0

```{r}


AWCD_cont <- 
  AWCD %>% 
  filter(Tox == 0) %>% 
  ungroup() %>% 
  group_by(Sampling, DIV, dhour) %>% 
  summarise( AWCD = median(AWCD)) %>% 
  spread(Sampling, AWCD) %>% 
  rename(Cont_Mid = Mid,
         Cont_End = End)
    


AWCD %>% 
  filter(Tox != 0) %>% 
  ungroup() %>% 
  spread(Sampling, AWCD) %>% 
  left_join(AWCD_cont) %>% 
  mutate(AWCD_inhib = 100 - 100 * (End / Cont_End) * (Cont_Mid / Mid)) %>% 
  #filter(DIV != 1) %>% 
  ggplot(., aes(x = ToxC, y = AWCD_inhib))+
  geom_point(shape = 21, colour = "darkblue")+
  geom_hline(aes(yintercept = 0), linetype = "dashed")+
  facet_grid(dhour ~ DIV, scales = "free")+
  scale_x_log10(breaks = signif(unique(ID$ToxC), 2))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = -45, hjust = 0))




filter(Bac_inhib, Cells == "CellsT") %>% 
ggplot(., aes(x = ToxC, y=prct_inhib))+
  facet_grid(DIV ~ sampling, labeller = label_wrap_gen(multi_line = F), scales = "free_y")+
  geom_point(position = position_jitter(width = 0.2), colour = "darkgreen")+
  geom_line(colour = "darkgreen")+
  geom_hline(yintercept = 0, linetype = "dashed")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45,hjust = 1, vjust = 1 ))+
  scale_x_log10(breaks = unique(Bac_inhib$ToxC))

filter(Bac_inhib, Cells == "CellsL") %>% 
ggplot(., aes(x = ToxC, y=prct_inhib))+
  facet_grid(DIV ~ sampling, labeller = label_wrap_gen(multi_line = F), scales = "free_y")+
  geom_point(position = position_jitter(width = 0.2), colour = "darkred")+
  geom_line(colour = "darkred")+
  geom_hline(yintercept = 0, linetype = "dashed")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45,hjust = 1, vjust = 1 ))+
  scale_x_log10(breaks = unique(Bac_inhib$ToxC))

```

